Investigating Fire Incidents Severity Levels in Toronto: Enhancing Resource Allocation for Fire Response

### 2024-12-11 paper/paper.qmd
Change graphs, include more details

# Data {#sec-data}

## Overview

@@

```{r}
#| label: fig-final-incident-type-distribution
#| fig-cap: Bar plot illustrating the distribution of final incident types across fire incidents. The plot presents the frequency of each incident type, with the count displayed on top of the bars. We can see that there is much more fire than explotion.
#| echo: false

ggplot(fire_data, aes(x = final_incident_type)) +
  geom_bar(fill = "purple") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), 
    plot.margin = margin(10, 10, 10, 10)  # Add space around the plot
  ) +
  scale_x_discrete(labels = function(x) sapply(x, function(y) paste(head(strsplit(y, " ")[[1]], 3), collapse = " "))) +
  xlab("Final Incident Type") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5)

@@

```{r}
#| label: fig-financial-loss-vs-casualties-heatmap
#| fig-cap: Heatmap illustrating the relationship between financial loss and casualty groups in fire incidents. The plot categorizes incidents based on financial loss and total casualties, with the color intensity representing the frequency of incidents in each combination of loss and casualty group. This provides insights into the severity of fire incidents, highlighting areas with higher financial losses and casualties. We can see that there are greater anount in 1k -10k region.
#| echo: false
#| warning: false

@@

ggplot(severity_table, aes(x = loss_group, y = casualty_group, fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    x = "Financial Loss Group",
    y = "Casualty Group",
    fill = "Incident Count"
  ) +
  theme_minimal()

@@

```{r}
#| label: fig-responding-apparatus-vs-log-dollar-loss
#| fig-cap: Scatter plot showing the relationship between the number of responding apparatus and the log-transformed financial loss in fire incidents. Each point represents a fire incident, with the x-axis displaying the log of the financial loss (with a 1 added for stability) and the y-axis showing the number of responding apparatus. There is a slight trend observed, indicating potential relationship.
#| echo: false

@@

# Model {#sec-model}

## Define Severity

@@

## Model Explnation

@@

@@

# Results {#sec-results}

## Logistic Regression Model Results

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-logisticresults-summary
#| fig-cap: "Table showing the summary of logistic regression model estimates for predicting fire severity. The table includes coefficients for each significant predictor, their standard errors, z-values, and associated p-values. Significant predictors are highlighted, with p-values less than 0.05 indicating statistical significance. The results provide insights into the relationship between 'Area of Origin' and the likelihood of high-severity fire incidents."

library(rstanarm)
library(modelsummary)
library(tidyverse)
library(dplyr)
library(caret)
library(xgboost)
library(kableExtra)

@@

# Print or work with the coefficients
kable(coefficients_table)
```

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-estimatedloss
#| tbl-cap: The table summarizes the distribution of estimated dollar losses from fire incidents.
library(knitr)

## 3. Estimated Dollar Loss (Numerical) ##
# Overall Table
est_loss_df <- as.data.frame(table(fire_data$estimated_dollar_loss))
colnames(est_loss_df) <- c("Estimated of Loss", "Frequency")
kable(summary(est_loss_df))
# Summary table
# Plot
# (No plot code provided in the original snippet)

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-responding-apparatus
#| fig-cap: This histogram illustrates the distribution of the number of responding apparatus for fire incidents. The x-axis represents the number of apparatus dispatched, while the y-axis shows the frequency of incidents. The date is right skewed, closer to 1.
## 7. Number of Responding Apparatus (Numerical) ##
ggplot(fire_data, aes(x = number_of_responding_apparatus)) +
  geom_histogram(bins = 30, fill = "cyan", color = "black") +
  theme_minimal() +
  xlab("Number of Responding Apparatus") +
  ylab("Frequency") +
  stat_bin(aes(label = ..count..), geom = "text", vjust = -0.5, size = 3)

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-responding-personnel-hist
#| fig-cap: This histogram shows the distribution of the number of responding personnel for fire incidents. The x-axis indicates the number of personnel dispatched, while the y-axis represents the frequency of such incidents. Similar dsitribution as @fig-responding-apparatus. 
# 8. Number of Responding Personnel (Numerical) ##
#Plot
ggplot(fire_data, aes(x = number_of_responding_personnel)) +
  geom_histogram(bins = 30, fill = "yellow", color = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("Number of Responding Personnel") +
  ylab("Frequency") +
  stat_bin(aes(label = ..count..), geom = "text", vjust = 0.5, size = 3)

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-possible-cause-full
#| tbl-cap: Comprehensive table showing all possible fire casues.
## 9. Possible Cause ##
#plot all that has frequency greater than 550
filtered_data <- fire_data |>
  count(possible_cause) |>
  filter(n > 550)
#Full Tabble
cause_table <- table(fire_data$possible_cause)
cause_table_df <- as.data.frame(cause_table) |>
  arrange(desc(Freq))
kable(cause_table_df, col.names = c("Possible Cause", "Frequency"))
```

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-fire-incidents-by-year
#| fig-cap: This bar chart shows the number of fire incidents by year, from 2011 to 2023.
## 12. tfs_alarm_time ## 
fire_data$year <- year(fire_data$tfs_alarm_time)
fire_data$month <- month(fire_data$tfs_alarm_time)
fire_data$month <- as.numeric(fire_data$month)

# Plot Number of Incidents by Year
ggplot(fire_data, aes(x = year)) +
  geom_bar(fill = "lightblue", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5, size = 3) + 
  labs(title = "Fire Incidents by Year", x = "Year", y = "Incident Count") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(fire_data$year), max(fire_data$year), by = 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-year-month-incident-count
#| fig-cap: Monthly Incident Count by Year (Excluding 2022 and 2023 becasue from prvious plot, we can see they don't contribute much information)

# from year plot, we can see that 2022 and 2023 does not provide much information so we can droped them.
fire_data_filtered <- fire_data |>
  filter(!is.na(year) & !is.na(month) & month >= 1 & month <= 12)
incident_count_by_year_month <- fire_data_filtered |>
  group_by(year, month) |>
  summarize(count = n(), .groups = 'drop')
incident_count_by_year_month_ex <- incident_count_by_year_month |>
  filter(!(year %in% c(2022, 2023)))
incident_count_by_year_month_ex$month <- factor(incident_count_by_year_month_ex$month, levels = 1:12)
ggplot(incident_count_by_year_month_ex, aes(x = month, y = count)) +
  geom_bar(stat = "identity", fill = "lightcoral", color = "black") +
  labs( x = "Month", y = "Incident Count") +
  scale_x_discrete(
    breaks = 1:12,
    labels = month.name 
  ) +
  facet_wrap(~year, scales = "fixed") +  # Fix the y-axis scale for comparison
  theme_minimal() +
  theme(
    strip.text.x = element_text(size = 10), 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1.5, size = 8), 
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12), 
    strip.background = element_rect(fill = "lightblue", color = "black"), 
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  )

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-firefighter-casulties
#| fig-cap: The Pie chart showing the distribution of firefighter casualties, rnaging from 0-3. With being the maximum amount of occurence.

## 14. Firefighter Casualties (Numerical) ##
casualties_table <- table(fire_data$tfs_firefighter_casualties)
casualties_df <- as.data.frame(casualties_table)
# Plot the pie chart
ggplot(casualties_df, aes(x = "", y = Freq, fill = factor(Var1))) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(theta = "y") +
  theme_void() +
  ggtitle("Distribution of Firefighter Casualties") +
  scale_fill_manual(values = c("pink", "lightblue", "lightgreen", "orange")) +
  labs(fill = "Casualties")
#table
#table(fire_data$tfs_firefighter_casualties)
```

@@

>>>>

# Data {#sec-data}
## Overview

@@

```{r}
#| label: fig-final-incident-type-distribution
#| fig-cap: Bar plot illustrating the distribution of final incident types across fire incidents. The plot presents the frequency of each incident type, with the count displayed on top of the bars. We can see that there is much more fire than explotion.
#| echo: false
#| fig.width: 3
#| fig.height: 2

ggplot(fire_data, aes(x = final_incident_type)) +
  geom_bar(fill = "purple") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1,size =6), 
    plot.margin = margin(10, 10, 10, 10)  # Add space around the plot
  ) +
  scale_x_discrete(labels = function(x) sapply(x, function(y) paste(head(strsplit(y, " ")[[1]], 3), collapse = " "))) +
  xlab("Final Incident Type") +
  ylab("Frequency") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0.5, size = 2)

@@

```{r}
#| label: fig-financial-loss-vs-casualties-heatmap
#| fig-cap: Heatmap illustrating the relationship between financial loss and casualty groups in fire incidents. The plot categorizes incidents based on financial loss and total casualties, with the color intensity representing the frequency of incidents in each combination of loss and casualty group. This provides insights into the severity of fire incidents, highlighting areas with higher financial losses and casualties. We can see that there are greater anount in 1k -10k region.
#| echo: false
#| warning: false
#| fig.width: 3
#| fig.height: 2

@@

ggplot(severity_table, aes(x = loss_group, y = casualty_group, fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(
    x = "Financial Loss Group",
    y = "Casualty Group",
    fill = "Incident Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 0.5,size =6)
  )

@@

```{r}
#| label: fig-responding-apparatus-vs-log-dollar-loss
#| fig-cap: Scatter plot showing the relationship between the number of responding apparatus and the log-transformed financial loss in fire incidents. Each point represents a fire incident, with the x-axis displaying the log of the financial loss (with a 1 added for stability) and the y-axis showing the number of responding apparatus. There is a slight trend observed, indicating potential relationship.
#| echo: false
#| fig.width: 3
#| fig.height: 2

@@

# Model {#sec-model}
## Define Severity

@@

## Model Explanation

@@

## Model Limitation
The logistic regression model has several limitations. First, it only considers predictors such as area of origin and ignition source, excluding potentially significant factors like weather conditions and building materials, which could affect fire severity. Additionally, the model assumes that the relationships between predictors and severity are linear, which may not fully capture the complexity of fire incidents. Most importantly, the categorization of severity is based on predetermined thresholds, which may not be universally applicable, as different contexts may require different severity definitions. Lastly, the model's predictive accuracy is influenced by the quality of the data.

Notably, there is a slight class imbalance in the dataset, high-severity incidents are underrepresented. In the training dataset, only 74 out of 10,237 records are classified as high severity. This imbalance can affect the model’s accuracy and its ability to predict high-severity incidents effectively.

@@

# Results {#sec-results}
## Logistic Regression Model Results

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-logisticresults-summary
#| fig-cap: "Table showing the summary of logistic regression model estimates for predicting fire severity. The table includes coefficients for each significant predictor, their standard errors, z-values, and associated p-values. Significant predictors are highlighted, with p-values less than 0.05 indicating statistical significance. The results provide insights into the relationship between 'Area of Origin' and the likelihood of high-severity fire incidents."
#| fig.align: left
#| fig.width: 6
#| fig.height: 4
library(rstanarm)
library(modelsummary)
library(tidyverse)
library(dplyr)
library(caret)
library(xgboost)
library(kableExtra)

@@

rownames(coefficients_table) <- gsub(
  "area_of_origin_grouped", "Area of Origin: ", 
  gsub("([a-z])([A-Z])", "\\1 \\2", rownames(coefficients_table)) # Add space before capital letters
)

# Print or work with the coefficients
kable(coefficients_table, digits = 2)

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-estimatedloss
#| tbl-cap: The table summarizes the distribution of estimated dollar losses from fire incidents.
#| fig.width: 3
#| fig.height: 2

library(knitr)

## 3. Estimated Dollar Loss (Numerical) ##
# Overall Table
est_loss_df <- as.data.frame(table(fire_data$estimated_dollar_loss))
colnames(est_loss_df) <- c("Estimated of Loss", "Frequency")
kable(summary(est_loss_df))
# Summary table
# Plot
# (No plot code provided in the original snippet)
```

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-responding-apparatus
#| fig-cap: This histogram illustrates the distribution of the number of responding apparatus for fire incidents. The x-axis represents the number of apparatus dispatched, while the y-axis shows the frequency of incidents. The date is right skewed, closer to 1.
#| fig.width: 3
#| fig.height: 2

## 7. Number of Responding Apparatus (Numerical) ##
ggplot(fire_data, aes(x = number_of_responding_apparatus)) +
  geom_histogram(bins = 30, fill = "cyan", color = "black") +
  theme_minimal() +
  xlab("Number of Responding Apparatus") +
  ylab("Frequency") +
  stat_bin(aes(label = ..count..), geom = "text", vjust = -0.5, size = 2)

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-responding-personnel-hist
#| fig-cap: This histogram shows the distribution of the number of responding personnel for fire incidents. The x-axis indicates the number of personnel dispatched, while the y-axis represents the frequency of such incidents. Similar dsitribution as @fig-responding-apparatus. 
#| fig.width: 3
#| fig.height: 2

# 8. Number of Responding Personnel (Numerical) ##
#Plot
ggplot(fire_data, aes(x = number_of_responding_personnel)) +
  geom_histogram(bins = 30, fill = "yellow", color = "black") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  xlab("Number of Responding Personnel") +
  ylab("Frequency") +
  stat_bin(aes(label = ..count..), geom = "text", vjust = 0.5, size = 2)

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-possible-cause-full
#| tbl-cap: A comprehensive table showing top 6 possible fire casues.
#| fig.width: 3
#| fig.height: 2

## 9. Possible Cause ##
#plot all that has frequency greater than 550
filtered_data <- fire_data |>
  count(possible_cause) |>
  filter(n > 550)
#Full Tabble
cause_table <- table(fire_data$possible_cause)
cause_table_df <- as.data.frame(cause_table) |>
  arrange(desc(Freq))
kable(head(cause_table_df), col.names = c("Possible Cause", "Frequency"))
```

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-fire-incidents-by-year
#| fig-cap: This bar chart shows the number of fire incidents by year, from 2011 to 2023.
#| fig.width: 3
#| fig.height: 2

## 12. tfs_alarm_time ## 
fire_data$year <- year(fire_data$tfs_alarm_time)
fire_data$month <- month(fire_data$tfs_alarm_time)
fire_data$month <- as.numeric(fire_data$month)

# Plot Number of Incidents by Year
ggplot(fire_data, aes(x = year)) +
  geom_bar(fill = "lightblue", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = 0, size = 2) + 
  labs(title = "Fire Incidents by Year", x = "Year", y = "Incident Count") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(fire_data$year), max(fire_data$year), by = 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-year-month-incident-count
#| fig-cap: Monthly Incident Count by Year (Excluding 2022 and 2023 becasue from prvious plot, we can see they don't contribute much information)
#| fig.width: 5
#| fig.height: 4

# from year plot, we can see that 2022 and 2023 does not provide much information so we can droped them.
fire_data_filtered <- fire_data |>
  filter(!is.na(year) & !is.na(month) & month >= 1 & month <= 12)
incident_count_by_year_month <- fire_data_filtered |>
  group_by(year, month) |>
  summarize(count = n(), .groups = 'drop')
incident_count_by_year_month_ex <- incident_count_by_year_month |>
  filter(!(year %in% c(2022, 2023)))
incident_count_by_year_month_ex$month <- factor(incident_count_by_year_month_ex$month, levels = 1:12)
ggplot(incident_count_by_year_month_ex, aes(x = month, y = count)) +
  geom_bar(stat = "identity", fill = "lightcoral", color = "black") +
  labs( x = "Month", y = "Incident Count") +
  scale_x_discrete(
    breaks = 1:12,
    labels = month.name 
  ) +
  facet_wrap(~year, scales = "fixed") +  # Fix the y-axis scale for comparison
  theme_minimal() +
  theme(
    strip.text.x = element_text(size = 6), 
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1.5, size = 4), 
    axis.text.y = element_text(angle = 45, hjust = 1, vjust = 1.5, size = 4),
    axis.title.x = element_text(size = 10), 
    axis.title.y = element_text(size = 10), 
    strip.background = element_rect(fill = "lightblue", color = "black"), 
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) 
```

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-firefighter-casulties
#| fig-cap: The Pie chart showing the distribution of firefighter casualties, rnaging from 0-3. With being the maximum amount of occurence.
#| fig.width: 3
#| fig.height: 2

## 14. Firefighter Casualties (Numerical) ##
casualties_table <- table(fire_data$tfs_firefighter_casualties)
casualties_df <- as.data.frame(casualties_table)
# Plot the pie chart
ggplot(casualties_df, aes(x = "", y = Freq, fill = factor(Var1))) +
  geom_bar(stat = "identity", width = 1, color = "black") +
  coord_polar(theta = "y") +
  theme_void() +
  ggtitle("Distribution of Firefighter Casualties") +
  scale_fill_manual(values = c("pink", "lightblue", "lightgreen", "orange")) +
  labs(fill = "Casualties")

#table
#table(fire_data$tfs_firefighter_casualties)
```

@@

# Idealized Survey Methodology and Data Collection for Fire Severity Study in Toronto
This study investigates the severity of fire incidents in Toronto to enhance resource allocation and optimize the effectiveness of the Toronto Fire Services (TFS). I work to understand key factors contributing to fire severity, including fire origin areas and ignition sources. I will also explore the effectiveness of public fire response systems, such as fire alarms sprinkler, smoke detection systems, to see how see thier presenace when fire incidence came and their effectiveness on reducsing fire. The methodology focuses on using observational data from fire incident records, public reports, and survey, and with statistical modeling to identify critical patterns and guide recommendations for TFS operations with understaingd of current fire infratcure presense.
